### 2025-10-25 thesis/implementation/index.tex
More impl content

The example selection is implemented using a multi-dimensional similarity
scoring system that identifies the most relevant samples that were previously
indexed for subsequent in-context learning. The component addresses the
fundamental challenge of selecting contextually appropriate examples from large
training corpus while balancing semantic relevance and structural compatibility.

@@

As this algorithm is aware of three distinct ways to measure sample similarity
(masked question, masked query, structural). The implementation uses
empirically optimized weights defined as constants: 70\% for semantic question
similarity and 30\% for SQL structural similarity. Additionally, the final
scoring combines 70\% sample-level similarity with 30\% schema-level
compatibility.

The selection algorithm yields a maximum of 32 candidates (\texttt{TOP\_K =
32}) to limit computational overhead while maintaining selection quality.
Vector similarity queries are performed using SQLite with the sqlite-vec
extension, achieving sub-second retrieval times for vector databases exceeding
2,000,000 samples.

@@

The \texttt{selection} algorithm implementation follows a two-stage approach,
split into initial candidate retrieval through vector search and subsequent
reweighting using the above described weights and a precomputed WWL distance
index from \texttt{natural-graphs}.

>>>>

The example selection system is implemented using a multi-dimensional similarity
scoring system that identifies the most relevant samples that were previously
indexed for subsequent in-context learning. The component addresses the
fundamental challenge of selecting contextually appropriate examples from large
training corpora whilst balancing semantic relevance and structural compatibility.

@@

As this algorithm is aware of three distinct ways to measure sample similarity
(masked question, masked query, structural). The implementation employs
empirically optimized weights defined as constants: 70\% for semantic question
similarity and 30\% for SQL structural similarity. Additionally, the final
scoring combines 70\% sample-level similarity with 30\% schema-level
compatibility.

The selection algorithm yields a maximum of 32 candidates (\texttt{TOP\_K =
32}) to limit computational overhead whilst maintaining selection quality.
Vector similarity queries are performed using SQLite with the \texttt{sqlite-vec}
extension, achieving sub-second retrieval times for vector databases exceeding
2,000,000 samples.

@@

The \texttt{selection} algorithm implementation follows a two-stage approach,
split into initial candidate retrieval through vector search (using the
\texttt{Vector} struct and SQLite) and subsequent reweighting using the above
described weights and a precomputed WWL distance index from
\texttt{natural-graphs}.

\subsubsection{Schema Subsetting System ($\phi$)}

The schema subsetting system is implemented in
\texttt{natural-inference::pipeline::subsetting} as
\texttt{SchemaSubsetter} struct. The core algorithm is the \texttt{optimize}
method (lines 21-47) which performs query validation to determine
which tables are crucial for the query candidates provided.

\paragraph{Query Validation}

Query validation employs a trial-and-error approach where for each query
candidate and table in the schema, an in-memory SQLite database
is created using \texttt{Connection::open\_in\_memory()}. Subsequently the
connection is initialized using the schema with all tables in the schema except
the current one.

Once the connection is ready, the \texttt{SchemaSubsetter} prepares every
query candidate against this reduced schema. If preparation fails, this
indicates the table removed is crucial for the query, thus it gets added to the
list of crucial tables. This process is repeated for all query-table
combinations to build a minimal schema containing only essential tables for the
execution of the set of query candidates provided to the subsetting algorithm.

\paragraph{Performance and Trade-off Characteristics}

As a new in-memory database for each table-query combination is created, this
leads to $O(queries \times tables)$ complexity. \textcolor{red}{For schemas
with XXX+ tables and multiple query candidates, this results in XXX-XXXms
processing time compared to 10-15ms for heuristic approaches.}

However, this execution-based approach provides superior correctness guarantees since heuristic
approaches cannot determine whether candidates will execute successfully in real
database environments. Using a real SQLite in memory instance both the correctness of
query candidates and the actually referenced set of tables can be known prior
to actual query execution.

\paragraph{Pipeline Integration}

The \texttt{SchemaSubsetter} is used prior to prompting the model using ICL to
ensure that the model context is used efficiently and attention is given to the
relevant parts of the schema. The \texttt{optimize} method returns a
\texttt{Schema} object containing only the tables identified as crucial through
execution testing which is in turn processed by the subsequent pipeline stages
like generation and refinement.

\subsubsection{Query Projection ($\pi$)}

The ICL module is implementing the query projection algorithm described in
section~\ref{design:projection-function}. It is implemented in the
\texttt{ICLGenerator} struct and wraps the \texttt{SqlModel} struct from
\texttt{natural-models}. Using llama.cpp it runs model inference using a
prompt optimized for in-context-learning with \textsc{OmniSQL}.

\paragraph{In-Context Learning}

The \texttt{ICLGenerator::generate} method implements few-shot prompting
with relevance filtering. Relevance filtering refers to removing all selected
samples with a similarity of less than 0.5 (this value was emperically derived
from evaluations). Thus only semantically or structurally similar samples are
actually provided to the model.

The prompt is constructed based on an adapted version of the \textsc{OmniSQL}
format \citep{OmniSQL}: task overview, sql schema, filtered examples with
similarity scores, explicit instructions, and the target question.
The biggest differentiation to the prompt of \citeauthor{OmniSQL} is the
example section including similarity. For the actual SQL query presentation the
code representation prompt format is used, inspired by DAIL-SQL
\citep{DAIL-SQL}. Every example includes the masked question, similarity score,
and formatted SQL query for clarity.

\begin{verbatim}
EXAMPLE OF THE PROMPT
\end{verbatim}

\paragraph{Prompt Engineering Strategy}

As outlined above the \textsc{OmniSQL} prompt was used as base for
\textsc{Natural}'s prompt together with a . \textsc{Natural} uses more explicit instructions
compared to DAIL-SQL. Key differentiations include the inclusion of similarity
scores to give the model the ability to weight the samples itself and

The prompt emphasizes precision and chain-of-thought reasoning. 
This aims to address common LLM issues with over-generating information or missing
query constraints, as well as the hallucination and accuracy concerns
identified in literature review section where LLMs generate plausible but
incorrect SQL queries.

\paragraph{Model Integration and Performance}

The \texttt{SqlModel} is wrapped around \texttt{llama-cpp-2} bindings and
loaded globally at startup to avoid a 30-60 second initialization times per
query. Using the \texttt{prompt} method from \texttt{natural-models}
tokenization, inference, and decoding with configurable PromptParams for
context size and generation limits is handled automatically.

\paragraph{Output Processing and Validation}

As \textsc{OmniSQL} was finetuned to output its thoughts and predictions using
markdown a markdown postprocessing module is needed, as well as a module to
identify whether a SQL query is syntactically valid.

To extract possible candidates all generated responses are post-processed
through a markdown parser (\texttt{pulldown-cmark}) which parses the model
output and looks for the code-block fence characters \texttt{\`\`\` }.

Subsequently all candidates are processed in reversed order and the first full
valid query is returned as potential candidate. The reversing of processing
order is needed as the model outputs it's thoughts top-to-bottom as a markdown
document with the most likely answer usually being output at the end. This
approach ensures that a returned query candidates are executable and
ensured to contain valid SQL. Thus \textsc{Natural} can aid the
difficulty of generating perfect SQL queries by acknowledging
limitations from large language models and implementing recovery and refinement
mechanisms in the subsequent pipeline flow.
